---
title: "P8105 Homework 2 — hy2951"
author: "Haochen Yu"
output:
  github_document:
    html_preview: false
editor_options:
  chunk_output_type: inline
---
```{r}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, fig.path="figs/")

dir.create("data", showWarnings = FALSE)
dir.create("figs", showWarnings = FALSE)
list.files("data")
```

```{r}
#libraries
library(tidyverse)
library(lubridate)
#pols-clean
pols <- read_csv("pols-month.csv", show_col_types = FALSE) |>
  separate(mon, into = c("year","month","day"), sep = "-", convert = TRUE) |>
  mutate(
    month = month(month, label = TRUE, abbr = FALSE),
    president = if_else(prez_gop == 1, "gop", "dem")
  ) |>
  select(year, month, president, everything(), -day, -prez_gop, -prez_dem) |>
  arrange(year, month)
#snp-clean
snp <- read_csv("snp.csv", show_col_types = FALSE) |>
  separate(date, into = c("m","d","y"), sep = "/", convert = TRUE) |>
  mutate(
    y = if_else(y < 50, 2000 + y, if_else(y < 100, 1900 + y, y)),
    month = month(m, label = TRUE, abbr = FALSE),
    year  = y
  ) |>
  transmute(year, month, close) |>
  arrange(year, month)
# unemp-tidy
unemployment <- read_csv("unemployment.csv", show_col_types = FALSE) |>
  pivot_longer(-Year, names_to = "mon", values_to = "unemployment_rate") |>
  mutate(
    mon   = str_sub(mon, 1, 3),
    mnum  = match(tolower(mon), tolower(month.abb)),
    month = month(mnum, label = TRUE, abbr = FALSE),
    year  = Year
  ) |>
  transmute(year, month, unemployment_rate) |>
  arrange(year, month)
#merge
pols_snp <- pols |> left_join(snp, by = c("year","month"))
p1_final <- pols_snp |> left_join(unemployment, by = c("year","month"))

dim_p1   <- dim(p1_final)
yr_min   <- min(p1_final$year, na.rm = TRUE)
yr_max   <- max(p1_final$year, na.rm = TRUE)

head(p1_final, 3)
```
In this problem, I cleaned and merged three monthly datasets from FiveThirtyEight:pols-month.csv（counts of national politicians by party, and the party of the sitting president (gop/dem).）snp.csv（S&P stock index closing values.）unemployment.csv（monthly U.S. unemployment rates.）
After cleaning, I merged them by year and month.
The final dataset has `r dim_p1[1]` rows and `r dim_p1[2]`columns, covering years from `r yr_min` to `r yr_max`. Key variables include the party of the president, counts of governors/senators/representatives, S&P closing values, and unemployment rates.

