---
title: "P8105 Homework 2 — hy2951"
author: "Haochen Yu"
output:
  github_document:
    html_preview: false
editor_options:
  chunk_output_type: inline
---
```{r}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, fig.path="figs/")

dir.create("data", showWarnings = FALSE)
dir.create("figs", showWarnings = FALSE)
list.files("data")
```

```{r}
#libraries
library(tidyverse)
library(lubridate)
#pols-clean
pols <- read_csv("pols-month.csv", show_col_types = FALSE) |>
  separate(mon, into = c("year","month","day"), sep = "-", convert = TRUE) |>
  mutate(
    month = month(month, label = TRUE, abbr = FALSE),
    president = if_else(prez_gop == 1, "gop", "dem")
  ) |>
  select(year, month, president, everything(), -day, -prez_gop, -prez_dem) |>
  arrange(year, month)
#snp-clean
snp <- read_csv("snp.csv", show_col_types = FALSE) |>
  separate(date, into = c("m","d","y"), sep = "/", convert = TRUE) |>
  mutate(
    y = if_else(y < 50, 2000 + y, if_else(y < 100, 1900 + y, y)),
    month = month(m, label = TRUE, abbr = FALSE),
    year  = y
  ) |>
  transmute(year, month, close) |>
  arrange(year, month)
# unemp-tidy
unemployment <- read_csv("unemployment.csv", show_col_types = FALSE) |>
  pivot_longer(-Year, names_to = "mon", values_to = "unemployment_rate") |>
  mutate(
    mon   = str_sub(mon, 1, 3),
    mnum  = match(tolower(mon), tolower(month.abb)),
    month = month(mnum, label = TRUE, abbr = FALSE),
    year  = Year
  ) |>
  transmute(year, month, unemployment_rate) |>
  arrange(year, month)
#merge
pols_snp <- pols |> left_join(snp, by = c("year","month"))
p1_final <- pols_snp |> left_join(unemployment, by = c("year","month"))

dim_p1   <- dim(p1_final)
yr_min   <- min(p1_final$year, na.rm = TRUE)
yr_max   <- max(p1_final$year, na.rm = TRUE)

head(p1_final, 3)
```
In this problem, I cleaned and merged three monthly datasets from FiveThirtyEight:pols-month.csv（counts of national politicians by party, and the party of the sitting president (gop/dem).）snp.csv（S&P stock index closing values.）unemployment.csv（monthly U.S. unemployment rates.）
After cleaning, I merged them by year and month.
The final dataset has `r dim_p1[1]` rows and `r dim_p1[2]`columns, covering years from `r yr_min` to `r yr_max`. Key variables include the party of the president, counts of governors/senators/representatives, S&P closing values, and unemployment rates.
```{r}
# ---- Problem 2 (final, for 2025 workbook) ----
library(tidyverse)   # wrangling
library(readxl)      # Excel I/O
library(lubridate)   # dates
library(knitr)       # preview table

# 1) Path (relative, as required)
path_tw <- "202509 Trash Wheel Collection Data.xlsx"
stopifnot(file.exists(path_tw))

# 2) Minimal name repair: lower_snake, no blank names, unique
repair_names <- function(nms) {
  nms[is.na(nms) | nms == ""] <- paste0("x", seq_len(sum(is.na(nms) | nms == "")))
  nms <- tolower(nms)
  nms <- gsub("[^a-z0-9]+", "_", nms)
  nms <- gsub("^_|_$", "", nms)
  make.unique(nms, sep = "_")
}

# 3) Reader for one sheet (meets the spec)
#    - skip note row(s); standardize names
#    - drop non-data rows
#    - round sports balls and cast to integer
#    - keep tidy columns and add wheel label
read_wheel <- function(sheet, label) {
  df <- read_excel(path_tw, sheet = sheet, skip = 1, .name_repair = "minimal")
  names(df) <- repair_names(names(df))

  # normalize key headers across versions
  nm <- names(df)
  dm <- nm[grepl("^dump|dumpster", nm, ignore.case = TRUE)]
  if (length(dm)) names(df)[match(dm[1], names(df))] <- "dumpster"

  dt <- nm[grepl("date", nm, ignore.case = TRUE)]
  if (length(dt)) names(df)[match(dt[1], names(df))] <- "date"

  # sports column can be "sports_balls" or just "sports"
  if ("sports" %in% names(df) && !"sports_balls" %in% names(df)) {
    names(df)[names(df) == "sports"] <- "sports_balls"
  }
  if (!"sports_balls" %in% names(df)) df$sports_balls <- NA_real_

  wt <- nm[grepl("weight.*ton", nm, ignore.case = TRUE)]
  if (length(wt) && wt[1] != "weight_tons") names(df)[match(wt[1], names(df))] <- "weight_tons"

  vol <- nm[grepl("volume.*yard", nm, ignore.case = TRUE)]
  if (length(vol) && vol[1] != "volume_cubic_yards") names(df)[match(vol[1], names(df))] <- "volume_cubic_yards"

  df |>
    mutate(
      date     = suppressWarnings(as_date(date)),
      dumpster = coalesce(as.integer(suppressWarnings(as.numeric(dumpster))), row_number())
    ) |>
    filter(!is.na(date)) |>
    mutate(
      year         = as.integer(year(date)),
      month        = month(date, label = TRUE, abbr = FALSE),
      sports_balls = as.integer(round(as.numeric(sports_balls))),
      wheel        = label
    ) |>
    select(any_of(c(
      "wheel","dumpster","date","year","month",
      "weight_tons","volume_cubic_yards",
      "plastic_bottles","polystyrene","cigarette_butts",
      "glass_bottles","plastic_bags","wrappers",
      "sports_balls","homes_powered"
    )))
}

# 4) Resolve the Gwynnda sheet name (e.g., "Gwynns Falls Trash Wheel")
sheet_gwyn <- readxl::excel_sheets(path_tw)[grepl("^Gwyn", readxl::excel_sheets(path_tw), ignore.case = TRUE)][1]

# 5) Read/clean three wheels
mr   <- read_wheel("Mr. Trash Wheel",       "Mr")
prof <- read_wheel("Professor Trash Wheel", "Professor")
gwyn <- read_wheel(sheet_gwyn,              "Gwynnda")

# 6) Combine tidy dataset and compute required summaries
trash <- bind_rows(mr, prof, gwyn) |> arrange(wheel, date)

n_obs  <- nrow(trash)
n_vars <- ncol(trash)

prof_total_weight <- trash |>
  filter(wheel == "Professor") |>
  summarize(total_weight_tons = sum(weight_tons, na.rm = TRUE)) |>
  pull()

gwyn_cigs_jun2022 <- trash |>
  filter(wheel == "Gwynnda", month(date) == 6, year(date) == 2022) |>
  summarize(total_cigs = sum(cigarette_butts, na.rm = TRUE)) |>
  pull()

# 7) Preview table (optional)
kable(head(trash, 5), caption = "Preview of combined Trash Wheel data (2025 workbook)")

```
In this problem, I imported and cleaned the 2025 Trash Wheel Collection Data for Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. I standardized column names, removed non-data rows, created year and month from dates, and converted the sports balls column to integers. After adding a wheel variable to identify each source, I combined the three datasets into one tidy dataset with `r n_obs` rows and `r n_vars` columns. From this data, Professor Trash Wheel collected a total of `r signif(prof_total_weight,6)` tons of trash, and Gwynnda collected `r gwyn_cigs_jun2022` cigarette butts in June 2022.